What is an Ancestral Recombination Graph?

### 2022-02-28 paper.tex

\section*{The Event ARG data structure}
% When did we move from the ARG is a process to the ARG is a data structure?
% What was the historical progression from one to the other?
Early work on ARG inference focused on the problem of
of inferring parameters of the
stochastic process, where the ancestry is regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}. These methods met with limited success
due to the overwhelmingly large and
horribly structured [Jere: what's the right way to say this?]
state space of ARGs. The focus subsequently shifted to
the more tractable---but still
NP-hard~\citep{wang2001perfect}---problem of computing
the minimum number of recombinations required
to explain an input dataset [CITATIONS], and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
The distinction between the ARG as a stochastic process
and the ARG as a data structure became less clear at this point,
with \citet{minichiello2006mapping} explicitly arguing for
the data structure interpretation.

% What do we mean by an EARG, informally?
The Griffiths ARG encoding is an elegant and mathematically
economical description of the coalescent with recombination on a single
chromosome, but it has significant drawbacks as the basis for a general purpose
storage and interchange format. To avoid confusion, we will refer to this
ARG encoding as an ``Event ARG'', or EARG.
The basic unit of an EARG are \emph{events} and correspond directly
to the events that
occur in the ARG stochastic process described in the previous section.
Each event corresponds to a node in the graph, and there are therefore
two types of node: common ancestor and recombination. As in the
stochastic process, edges represent lineages, and those lineages
are modified by the events that they pass through.

% What do we mean by an EARG, formally?
% NOTE we're using the natural numbers but are using 0 as a node in the
% examples. Probably want to clear this up a bit.
Formally, we can define an EARG as a tuple $(e, \sigma)$, where $e$
defines a directed acyclic graph where $e_j = (c, p)$ describes
an edge between
child node $c$ and parent $p$ ($c, p \in \mathbb{N}$).
Additionally,
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints, such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \infty$ otherwise. (The use of $\infty$ to denote
non-recombination nodes is arbitrary, but useful when we extract
% NOTE: need to be careful about using "local" trees if we're also
% distinguishing resolved and non-resolved GARGs as being "locally"
% defined or not. One is topologically local and the other is local
% along the genome.
local trees; see below.)
For each recombination node $u$ we must
have exactly two edges in $E$ such that $u$ is the child.
Note that this description captures only the
graph topology: if we also wish to know the times of events we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.
An example is given in Fig.~\ref{fig-arg-data-structure}A.

@@

\end{tikzpicture}
\caption{\label{fig-arg-data-structure}
Node vs edge breakpoint annotations.
The classical approach described by Griffiths
equates graph nodes with events in the stochastic process,
and associates recombination breakpoint information with \emph{nodes} (left).
This encoding
is limited in the patterns of inheritance that can be described
and requires that edges be ordered to avoid ambiguity.
[TODO clarify that the function is the same things as having two
types of nodes]
The alternative is to associate breakpoint annotations
with \emph{edges} (right).
Associating the interval of genome inherited by the child node
from the parent with the corresponding edge allows
any pattern genetic inheritance to be described and does
not require edges to be ordered.
}
\end{figure}

@@

The ordering of edges is a vital element of the EARG encoding.
This is because there is a fundamental ambiguity involved
in associating recombination breakpoints with \emph{nodes}
rather than edges,
and the only way we can break this symmetry is to assign
some meaning to the order in which the parents are listed
(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
parents of a recombination event). When we want to recover
the local tree at a position $x$ along the genome,
we traverse the graph upwards from the leaves. At a particular
node $u$, if it has one parent we are at a common ancestor
node and we simply follow that parent. If we are at a
recombination node, $u$ has two parents; if
$x$ is less than the breakpoint $\sigma(u)$ we follow
the first edge, and otherwise follow the second edge.
This ordering requirement, which although straightforward
to describe, has some drawbacks in practice. For example,
% Surely the meaning is clear from the context here and we don't need
% qualify the meaning of "ARG"
simulating an ARG in this representation is
complicated by the fact that the
first event to occur may hit the lineage carrying the ancestry
to the right of the breakpoint rather
than the left (and so we cannot emit edges as they are generated).
Such issues can be worked around, of course,
but depending on the ordering of otherwise indistinguishable
objects is generally problematic. In practice, several
methods
explicitly associate information with the outbound edges
of a recombination event
to resolve the problem~\citep{lyngso2005minimum,ignatieva2021kwarg}.

@@

The first type of simplification that we can perform is to remove
graph topology that is invisible to the samples. The most well
known example of such topology is a so-called
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g. $K,N,O$ and $R$ in
Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or ancestral genomes,
there is no information in this topology and the diamond can b
e replaced by a single edge. More generally, any
subgraph which is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by a single edge. This definition includes the case
of a single node that has one inbound and one outbound edge.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph-level simplification.

>>>>

\section*{The Event ARG data structure}
% When did we move from the ARG is a process to the ARG is a data structure?
% What was the historical progression from one to the other?
Early work on ARG inference focused on the problem of
of inferring parameters of the
stochastic process, where the ancestry is regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}.
% FIXME: need some proper statistical language here to describe why these
% methods didn't work so well.
These methods met with limited success
due to the overwhelmingly large and
horribly structured [Jere: what's the right way to say this?] state space of ARGs.
The focus subsequently shifted to the more
tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
the minimum number of recombinations required
to explain an input dataset [CITATIONS], and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
The distinction between the ARG as a stochastic process
and the ARG as a \emph{data structure}~\citep{minichiello2006mapping}
became important at this point,
since parsimonious ARGs have no immediate connection to
the stochastic process. Indeed, the ARG instances produced by
the coalescent with recombination are usually highly unparsimonious,
since they contain many invisible recombinations
(although those produced under the SMC may be parsimonious, depending
on the pattern of mutations that is present).

% What do we mean by an EARG, informally?
The Griffiths ARG encoding is an elegant and mathematically
economical description of the coalescent with recombination,
but it has significant drawbacks as the basis for a general purpose
storage and interchange format. To avoid confusion, we will refer to this
ARG encoding as an ``Event ARG'', or EARG.
The fundamental units of an EARG are the \emph{events} that correspond directly
to those that
occur in the ARG stochastic process described in the previous section.
Each event corresponds to a node in the graph, and there are therefore
two types of node: common ancestor and recombination. As in the
stochastic process, edges represent lineages, and those lineages
are modified by the events that they pass through.

@@

\end{tikzpicture}
\caption{\label{fig-arg-data-structure}
Node vs edge breakpoint annotations.
The classical approach described by Griffiths
equates graph nodes with events in the stochastic process,
and associates recombination breakpoint information with \emph{nodes} (left).
This encoding
is limited in the patterns of inheritance that can be described
and requires that edges be ordered to avoid ambiguity.
The alternative is to associate breakpoint annotations
with \emph{edges} (right).
Associating the interval of genome inherited by the child node
from the parent with the corresponding edge allows
any pattern genetic inheritance to be described and does
not require edges to be ordered.
}
\end{figure}

% What do we mean by an EARG, formally?
% NOTE we're using the natural numbers but are using 0 as a node in the
% examples. Probably want to clear this up a bit.
Formally, we can define an EARG as a tuple $(e, \sigma)$, where $e$
is an ordered list of edges defining a directed acyclic graph and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints,
such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \varnothing$ otherwise.
(Note that this formulation is equivalent to stipulating two
different ``types'' of node and only storing breakpoints
with recombination nodes).
Each edge $e_j = (c, p)$ describes an edge between
child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
For each recombination node $u$ we must
have exactly two edges in $e$ such that $u$ is the child.
Note that this description captures only the
graph topology: if we also wish to know the times of events we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.
An example is given in Fig.~\ref{fig-arg-data-structure}A.

@@

The ordering of edges is a vital element of the EARG encoding.
This is because there is a fundamental ambiguity involved
in associating recombination breakpoints with \emph{nodes}
rather than edges,
and the only way we can break this symmetry is to assign
some meaning to the order in which the parents are listed
(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
parents of a recombination event). This is usually described
in terms of the process in which we extract the local trees
along the genome from an EARG.
To recover
the tree at genome coordinate $x$,
we traverse the graph upwards from the leaves. At a particular
node $u$, if it has one parent we are at a common ancestor
node and we simply follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint $\sigma(u)$ we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
This ordering requirement, which is straightforward
to describe, has some practical drawbacks. For example,
% Surely the meaning is clear from the context here and we don't need
% qualify the meaning of "ARG"
simulating an ARG in this representation is
complicated by the fact that the
first event to occur may hit the lineage carrying the ancestry
to the right of the breakpoint rather
than the left (and so we cannot emit edges as they are generated).
Such issues can be worked around, of course,
but depending on the ordering of otherwise indistinguishable
objects is generally problematic. In practice, several
methods
explicitly associate information with the outbound edges
of a recombination event
to resolve the problem~\citep{lyngso2005minimum,ignatieva2021kwarg}.

@@

The first type of simplification that we can perform is to remove
graph topology that is invisible to the samples. The most well
known example of such topology is a so-called
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g. $K,N,O$ and $R$ in
Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or ancestral genomes,
there is no information in this topology and the diamond can b
e replaced by a single edge. More generally, any
subgraph which is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by a single edge. This definition includes the case
of a single node that has one inbound and one outbound edge.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph topology simplification.
